---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# idtools

<!-- badges: start -->
<!-- badges: end -->

`idtools` is an R package for extracting sample identifiers from filenames at the
North West Genomic Laboratory Hub in Manchester, United Kingdom. 
It is intended to be used for validation or development projects, where key 
information, such as the DNA lab number, worksheet number and replicate suffix, 
are concatenated within the filename. 

The name is a pun on the concept of "tidy tools" from the ["Tidy Tools Manifesto"](https://tidyverse.tidyverse.org/articles/manifesto.html) by Hadley
Wickham.

## Installation

You can install the development version of idtools from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("joe-m-shaw/labtools")
```

## Example

You can use `idtools` to extract sample identifiers stored in
filenames.

```{r}

library(idtools)

filename <- "WS123456_12345678a"

extract_worksheet(filename)

extract_labno(filename)

extract_suffix(filename)

```

## Information Governance

**No patient identifiable information should be included in this
repository.**

If you spot some, please let me know.

For the purpose of testing functions, I have used generic examples for
worksheet (WS123456) and lab number (12345678) values.

Where examples of patient names are required, I have used character
names from novels by Leo Tolstoy (Anna Karenina, Pierre Bezukhov etc).
