---
title: "Data analysis with idtools"
vignette: >
  %\VignetteIndexEntry{Data analysis with idtools}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: FALSE
library(idtools)
library(purrr)
library(readr)
library(ggplot2)
library(readxl)
```

`idtools` is designed to be used at the North West Genomic Laboratory Hub for 
projects involving data analysis of multiple results files. 
This vignette goes through an example of a common scenario where `idtools` can 
help.

# Lots of files with long filenames

A common scenario faced in genetic test development is where there are multiple files
of results, and the identifiers for each sample are included in long filenames.

Here is an example with 5 .csv files which contain gene dosage results for
fictional patients. All these files are saved in a folder called "data".

```{r}
#| label: csv-filepaths
#| include: TRUE

csv_files <- list.files(path = "data/",
           pattern = ".*.csv",
           full.names = TRUE)

csv_files

```

Each .csv file contains the dosage (log2 ratio) results for 5 genes of interest.

```{r}
#| echo: FALSE
#| label: tbl-csv-data-example
#| tbl-cap: "Contents of a single .csv file"

csv_data_example <- readr::read_csv(csv_files[1],
                                    show_col_types = FALSE)

knitr::kable(csv_data_example)

```

In order to analyse the data, two steps need to be completed:

1. The results from each individual file should be collated together in one
dataframe

2. The sample identifiers from each filename should be added to the 
dataframe as separate columns.

## 1. Collating the results with `purrr` and `readr`

Collating the file contents can be achieved
by applying the `read_csv` function from `readr` to all the files, using
the `map` function from `purrr`.
Specifying `id="filepath"` in the call to `read_csv` adds the full filepath
as a new column to the contents of each file.

```{r}
#| include: TRUE
#| label: collate-csv-files

csv_data <- csv_files |> 
  purrr::map(\(csv_files) readr::read_csv(csv_files,
                            id = "filepath",
                            show_col_types = FALSE)) |> 
  purrr::list_rbind() 

```

This creates a dataframe containing the results from all the csv files.

```{r}
#| label: tbl-collated-csv-data
#| echo: FALSE
#| tbl-cap: "Collated gene dosage data with filepaths"

knitr::kable(csv_data)

```

## 2. Adding sample identifiers with `idtools`

The `mutate_ids` function from `idtools` can then be applied to the new 
"filepath" column of the collated data.
This function extracts the DNA lab number (labno), replicate suffix and worksheet
number from the filepath, and adds each identifier as a new column.

```{r}
#| include: TRUE
#| label: mutate-ids

csv_data_with_names <- csv_data |> 
  idtools::mutate_ids(id_col = filepath)

```

```{r}
#| echo: FALSE
#| label: tbl-collated-data-with-names
#| tbl-cap: "Collated gene dosage data with separated identifiers"

knitr::kable(csv_data_with_names)

```

## Analysing the collated data

Now that the data is all together in a single dataframe, including the sample
identifiers, it can be more easily analysed.

For example, the data can be quickly filtered to return only the results for 
the *EGFR* gene.

```{r}
#| echo: TRUE
#| label: tbl-csv-data-gene
#| tbl-cap: "Results for EGFR gene"

knitr::kable(csv_data_with_names |> 
  dplyr::filter(gene == "EGFR"))

```

The data can also be filtered to show the results for the 3 different replicates 
of sample 12345678.

```{r}
#| echo: TRUE
#| label: tbl-csv-data-labno
#| tbl-cap: "Results for sample 12345678"

knitr::kable(csv_data_with_names |> 
  dplyr::filter(labno == "12345678"))

```

And the `ggplot2` package from the tidyverse can be used to visualise all the
results.

```{r}
#| echo: FALSE
#| label: plot

ggplot(csv_data_with_names, aes(x = gene, y = log2r)) +
  geom_jitter(shape = 21, size = 3,
              alpha = 0.6, width = 0.1, aes(fill = suffix)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~labno) +
  labs(x = "Gene", y = "Log2 ratio",
       title = "Collated csv file results",
       subtitle = "Facetted by lab number",
       fill = "Suffix")

```

## Reading different file types

The same data folder also contains 5 Excel files of quality control information.
However, the filename format is different: the identifiers are in a different
order to the csv files, and hyphens are used instead of underscores.

```{r}
#| label: xlsx-files
#| include: TRUE

xlsx_files <- list.files(path = "data/",
           pattern = ".*.xlsx",
           full.names = TRUE)

xlsx_files

csv_files

```

Each file contains the "noise" metric for each sample replicate.

```{r}
#| label: tbl-example-xlsx-file
#| echo: FALSE
#| tbl-cap: "Contents of a single .xlsx file"

xlsx_data_example <- readxl::read_excel(xlsx_files[1])

knitr::kable(xlsx_data_example)

```

Similarly to before, the contents of the xlsx files can be collated together
using the `readxl` package.
The `read_excel` function from `readxl` does not include an option to add the
filepath as a separate column, so an alternative approach is to use
`set_names` and the "names_to" argument in `list_rbind` to achieve the same
result.

```{r}
#| label: collate-xlsx-files
#| include: TRUE

xlsx_data <- xlsx_files |> 
  purrr::map(\(xlsx_files) readxl::read_excel(xlsx_files)) |> 
  rlang::set_names(xlsx_files) |> 
  purrr::list_rbind(names_to = "filepath") 

```

```{r}
#| label: tbl-collated-xlsx-data
#| echo: FALSE
#| tbl-cap: "Collated quality metric data with filepaths"

knitr::kable(xlsx_data)

```

Then, just as with the csv files, the identifiers can be extracted using
`mutate_ids`.
Importantly, this process still works even though the filename format has changed.

```{r}
#| label: mutate-ids-excel
#| include: TRUE

xlsx_data_with_names <- xlsx_data |> 
  idtools::mutate_ids(id_col = filepath)

```

```{r}
#| label: tbl-xlsx-data-with-names
#| echo: FALSE
#| tbl-cap: "Collated quality metric data with separate identifiers"

knitr::kable(xlsx_data_with_names)

```

