---
title: "Data analysis with idtools"
vignette: >
  %\VignetteIndexEntry{Data analysis with idtools}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(idtools)
library(purrr)
library(readr)
library(ggplot2)
library(readxl)
```

# Development projects, data analysis and the tidyverse

Developing new genetic tests involves working with data.

This means collating data from multiple worksheets, samples and replicates.

It also usually involves combining data which has been stored in different
file types (.csv, .xlsx, .vcf etc).

This is called ["Exploratory Data Analysis" (EDA)](https://r4ds.hadley.nz/EDA.html)
and 

# A common scenario

A common scenario faced in test development is where there are multiple files
of results, and the identifiers for each sample are included in the filename.

Here is an example with 5 csv which contain results for
fictional patients. All these files are saved in a folder called "data".

```{r}
#| label: filepaths
#| echo: FALSE

csv_files <- list.files(path = "data/",
           pattern = ".*.csv",
           full.names = TRUE)

csv_files

```

Each csv file contains the dosage (log2 ratio) results for 5 genes of interest.

```{r}
#| echo: FALSE

readr::read_csv(csv_files[1])

```

In order to analyse the data, two steps need to be completed:

- The results from each individual file should be collated together in one
dataframe

- The sample identifiers from each filename should be added to the 
dataframe as separate columns.

## Collating the results

Collating the file contents can be achieved
by applying the `read_csv` function from `readr` to all the files, using
the `map` function from `purrr`.
Specifying `id="filepath"` in the call to `read_csv` adds the full filepath
as a new column to the contents of each file.

```{r}
#| include: TRUE
#| label: collate-csv-files

csv_data <- csv_files |> 
  purrr::map(\(csv_files) readr::read_csv(csv_files,
                            id = "filepath",
                            show_col_types = FALSE)) |> 
  purrr::list_rbind() 

csv_data

```

## Adding sample identifiers

The `mutate_ids` function from `idtools` can then be applied to the new 
"filepath" column of the collated data.
This function extracts the DNA lab number (labno), replicate suffix and worksheet
number from the filepath, and adds each identifier as a new column.

```{r}
#| include: TRUE
#| label: mutate-ids

csv_data_with_names <- csv_data |> 
  idtools::mutate_ids(id_col = filepath)

csv_data_with_names

```

## Analysing the collated data

Now that the data is all together in a single dataframe, including the sample
identifiers, it can be more easily analysed.

For example, the data can be quickly filtered to return only the results for 
the *EGFR* gene.

```{r}
#| include: TRUE

csv_data_with_names |> 
  dplyr::filter(gene == "EGFR")

```

The data can also be filtered to show the results for the 3 different replicates 
of sample 12345678.

```{r}
#| include: TRUE

csv_data_with_names |> 
  dplyr::filter(labno == "12345678")

```

And the `ggplot2` package from the tidyverse can be used to visualise all the
results.

```{r}
#| echo: FALSE

ggplot(csv_data_with_names, aes(x = gene, y = log2r)) +
  geom_jitter(shape = 21, size = 3,
              alpha = 0.6, width = 0.1, aes(fill = suffix)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~labno) +
  labs(x = "Gene", y = "Log2 ratio",
       title = "Collated csv file results",
       subtitle = "Facetted by lab number",
       fill = "Suffix")

```

## Reading different file types

The same data folder also contains 5 Excel files
containing the "noise" metric for each sample replicate.
However, the filename format is different: the identifiers are in a different
order and hyphens are used instead of underscores.

```{r}
#| label: xlsx-files

xlsx_files <- list.files(path = "data/",
           pattern = ".*.xlsx",
           full.names = TRUE)

xlsx_files

readxl::read_excel(xlsx_files[1])

```

Similarly to before, the contents of the xlsx files can be collated together
using the `readxl` package.
The `read_excel` function from `readxl` does not include an option to add the
filepath as a separate column, so an alternative approach is to use
`set_names` and the "names_to" argument in `list_rbind` to achieve the same
result.

```{r}
#| label: collate-xlsx-files

xlsx_data <- xlsx_files |> 
  purrr::map(\(xlsx_files) readxl::read_excel(xlsx_files)) |> 
  rlang::set_names(xlsx_files) |> 
  purrr::list_rbind(names_to = "filename") |> 
  idtools::mutate_ids() 

xlsx_data

```

Then, just as with the csv files, the identifiers can be extracted using
`mutate_ids`, which still works even though the filename format has changed.

```{r}
#| label: mutate-ids-excel

xlsx_data_with_names <- xlsx_data |> 
  idtools::mutate_ids()

xlsx_data_with_names

```
