---
title: "The scope of idtools"
vignette: >
  %\VignetteIndexEntry{The scope of idtools}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: FALSE
library(idtools)
```

`idtools` is intended to work with tidyverse packages like `purrr` and `readr`
to make manipulating lab identifiers easier. A few key design choices of 
`idtools` are summarised below.

# Why is there no `extract_patient_name` function?

There is no function for extracting patient names in `idtools` for the following reasons:

- **Patient names are not always included in filenames.** Patient names may be
included in filenames if analysis is performed using on-prem servers, but if
data is analysed using cloud computing
then patient names are removed. This means datasets can include 
a mixture of files with and without patient names depending on how they were processed.

- **Not all samples are from people.** Some samples are control samples, such as mixtures 
of different DNA samples or cell lines. These samples may be given a "name" or they
may not, and the name may or may not follow the standard format of "forename surname".
Example: "HorizonPositiveControl".

- **Some samples may be for research.** Patient samples collected for research will 
be anonymised, and the research identifier may be included in place of the sample
name. The research identifier may also include numbers or other non-letter characters.
Example: "study0001"

These reasons mean that designing a regular expression to identify a patient name
in a filename is complicated.
One option is to use the position of the name in the string relative to more 
consistent identifiers, like lab number and worksheet number, but this is then 
inflexible when the order of identifiers changes.

A better approach is to get patient names from databases such as iGene or
DNA Database.

# Why is there no `read_with_filename` function?

Given that `idtools` is based around extracting identifiers from filenames, an
earlier version included a generic function to read files as dataframes and 
then add on a new column containing the filename. The prototype function is
shown below, and the idea was that it could be combined with multiple 
`read` functions like `readr::read_csv`, `readxl::read_excel` and 
`vcfR::read.vcfR`, and then piped into `mutate_ids`.

```{r}
#| include: TRUE
#| label: read-with-filename

read_with_filename <- function(filepath, read_fn) {

  output <- read_fn(filepath, ...) |>
    dplyr::mutate(filepath = filepath,
                  filename = basename(filepath))

  return(output)

}

```

However, there are better options for adding the filepath as a new column
when reading files. For `readr::read_csv` there is the "id" argument, and
for other functions there is the option of using `rlang::set_names` and
`purrr:list_rbind`. 
Both these options are shown in more detail in the
"Data analysis with idtools" vignette.
